---
title: "Using colocr"
author: "Mahmoud Shaaban"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# colocr

An R package for conducting co-localization analysis.

Install package from github

```{r, eval=FALSE}
devtools::install_github('MahShaaban/colocr')
```


## Getting started

To get started, load the required packages and the images.
Then, apply the appropriate parameters for choosing the regions of interest
using the `parameter_choose`. Finally, check the appropriatness of the 
parameters by highlighting the ROIs on the image.

```{r}
# load libraries
library(imager)
library(colocr)

# load images
fl <- system.file('extdata', 'Image0001_.jpg', package = 'colocr')
img <- load.image(fl)
img.g <- grayscale(img)

# choose parameters
px <- parameter_choose(img.g, threshold = 90)

# highlight chosen region of interest
plot(img)
highlight(px)
```

The same can be acheived interactively using an accompanying **shiny** app.
To launch the app run.

```{r, eval=FALSE}
run_app()
```

The reset of the anlysis depends on the particular kind of images. Now, `colocr`
implements two simple colocalizations statistics; Pearson's Coefficeint Correlation (PCC) and the Manders Overlap Coefficient (POC).

Try the shiny app [here](https://mahshaaban.shinyapps.io/colocr_app/)
